---
title: "Lab_9_Wakerley"
author: "Kyle Wakerley"
format: html
editor: visual
embed-resources: true
---

## Lab 9

```{r, warning=FALSE}

# install.packages(c("RSQLite", "DBI"))

library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Download tables
doctor <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/doctor.csv")
visit <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/visit.csv")
patient <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/patient.csv")
payment <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/payment.csv")

# Copy data.frames to database
dbWriteTable(con, "doctor", doctor)
dbWriteTable(con, "visit", visit)
dbWriteTable(con, "patient", patient)
dbWriteTable(con, "payment", payment)
```

```{r}
dbListTables(con)
```

## Exercise 1

```{sql connection=con}
SELECT doctor_id, first_name, last_name
FROM doctor
ORDER by last_name, first_name
```

The above code chunk retrieves the doctor ID, first name and last name for all doctors using the doctor table.

## Exercise 2

```{sql connection = con}
SELECT doctor_id, first_name, last_name
FROM doctor
WHERE last_name IN ('WILLIAMS', 'DAVIS')
```

There are 6 doctors with a last name of "Davis" or "Williams"

## Exercise 3

```{sql connection = con}
SELECT DISTINCT patient_id
FROM visit
WHERE date(visit_start_date) = '2005-07-05'
```

There are 27 patients who started a hospital visit on July 5th 2005.

### Exercise 4.1

```{sql connection = con}
SELECT *
FROM payment
WHERE amount IN (199, 799, 999)
```

The query resulted in 111 matched for payment amounts of exactly 199, 799, and 999.

### Exercise 4.2

```{sql connection = con}
SELECT *
FROM payment
WHERE amount > 500
```

There are 266 rows where the payment amount was greater than 500

### Exercise 4.3

```{sql connection = con}
SELECT *
FROM payment
WHERE amount > 500 AND amount < 800
```

There are 204 instances of payments greater than 500 and less than 800

## Exercise 5

```{sql connection = con}
SELECT payment_id, amount
FROM payment
  INNER JOIN patient ON payment.patient_id = patient.patient_id
WHERE last_name = "DAVIS"
```

This returned 3 payments where the last name was "Davis"

## Exercise 6

### Exercise 6.1

```{sql connection=con}
SELECT COUNT(*)
FROM visit
```

There are 16044 visits

### Exercise 6.2

```{sql connection = con}
SELECT COUNT(*), patient_id
FROM visit
GROUP BY patient_id
```

Above are the counts for each patient

### Exercise 6.3

```{sql connection = con}
SELECT COUNT(*), patient_id
FROM visit
GROUP BY patient_id
ORDER BY COUNT(*) DESC
```

Now we ordered it in decending order.

### Exercise 6.4

```{sql connection = con}
SELECT COUNT(*), patient_id
FROM visit
GROUP BY patient_id
HAVING COUNT(*) > 40
ORDER BY COUNT(*) DESC
```
Filtered by observations above 40

## Exercise 7

```{sql connection = con}
SELECT MAX(amount), MIN(amount), AVG(amount), SUM(amount)
FROM payment
```
Calculated summary stats. 

### Exercise 7.1

```{sql connection = con}
SELECT  patient_id, MAX(amount), MIN(amount), AVG(amount), SUM(amount)
FROM payment
GROUP BY patient_id
```

Grouped by patient!

### Exercise 7.2

```{sql connection = con}
SELECT  patient_id, MAX(amount), MIN(amount), AVG(amount), SUM(amount)
FROM payment
GROUP BY patient_id
HAVING COUNT(*) > 5
```

Only calculated summary stats for patients who had more than 5 observations.

### Exercise 8.1

```{sql connection = con}
SELECT AVG(julianday(visit_end_date) - julianday(visit_start_date)), doctor.doctor_id
FROM visit
JOIN doctor ON visit.doctor_id = doctor.doctor_id
GROUP BY doctor.doctor_id
```


### Exercise 8.2

```{sql connection = con}
SELECT CONCAT(doctor.first_name, " " , doctor.last_name, " " ,  "Had an Average Visit Time of" , " ", AVG(julianday(visit_end_date) - julianday(visit_start_date)))
FROM visit
JOIN doctor ON visit.doctor_id = doctor.doctor_id
GROUP BY doctor.doctor_id
ORDER BY AVG(julianday(visit_end_date) - julianday(visit_start_date)) DESC LIMIT 5;
```
```{sql connection = con}
SELECT CONCAT(doctor.first_name, " " , doctor.last_name, " " ,  "Had an Average Visit Time of" , " ", AVG(julianday(visit_end_date) - julianday(visit_start_date)))
FROM visit
JOIN doctor ON visit.doctor_id = doctor.doctor_id
GROUP BY doctor.doctor_id
ORDER BY AVG(julianday(visit_end_date) - julianday(visit_start_date)) ASC LIMIT 5;
```
```{r}
dbDisconnect(con)
```

