---
title: "Lab_8_Wakerley"
format: html
editor: visual
embed-resources: true
---

```{r}
library(microbenchmark)
```

## Problem 1: Vectorization

```{r}
fun1 <- function(n = 100, k = 4, lambda = 4) {
  x <- NULL
  
  for (i in 1:n){
    x <- rbind(x, rpois(k, lambda))    
  }
  
  return(x)
}

fun1alt <- function(n = 100, k = 4, lambda = 4) {
  matrix(rpois(n * k, lambda), nrow = n, ncol = k)
}

x1 = fun1()
x2 = fun1alt()

hist(as.vector(x1), main = "Poisson Distribution from fun1",
     xlab = "Value", col = "skyblue")

hist(as.vector(x2), main = "Poisson Distribution from fun1alt", xlab = "Value", col = "tomato")

microbenchmark(fun1(), fun1alt())

```

We can see that function 1 completes in an average of 236 microseconds, compared to only 15.4 with my alternate function 1. This is because rbind is very slow. Using vectorization (computing all the value at once) speeds the process up tremendously. 

### Part 2: Max of Each Column

```{r}
# Data Generating Process (10 x 10,000 matrix)
set.seed(1234)
x <- matrix(rnorm(1e4), nrow=10)

# Find each column's max value
fun2 <- function(x) {
  apply(x, 2, max)
}

fun2alt = function(x) {
  y = ncol(x)
  out = numeric(y)
  
  for (k in 1:y) {
    out[k] <- max(x[, k])
  }
  out
}

microbenchmark(fun2(x), fun2alt(x))

x11 = fun2(x)
x22 = fun2alt(x)

sum(x11-x22) #Sums to zero, all entries are the same
```

I checked that x11 and x22 produce the same results by summing the two together vector wise. The resulting sum i zero, confirming that the two outputs were the same. My function 2 appears to run slightly faster, with a mean time of 1140.911 microseconds compared to function 2 which runs at an average of 1469.913 microseconds. 

## Problem 2: Parallelization

```{r}
library(parallel)


my_boot <- function(dat, stat, R, ncpus = 2L) {
  cl <- makeCluster(ncpus)
  
  n <- nrow(dat)
  idx <- matrix(sample.int(n, n*R, TRUE), nrow=n, ncol=R)
  
  clusterExport(cl, c("dat", "idx", "stat"), envir = environment())
  
  ans <- parLapply(cl, seq_len(R), function(i) {
    stat(dat[idx[,i], , drop=FALSE])
  })
  
  # Converting the list into a matrix
  ans <- do.call(rbind, ans)

  return(ans)
}

# Bootstrap of an OLS
my_stat <- function(d) coef(lm(y ~ x, data=d))

# DATA SIM
set.seed(1)
n <- 500; R <- 1e4

x <- cbind(rnorm(n)); y <- x*5 + rnorm(n)

# Checking if we get something similar as lm
ans0 <- confint(lm(y~x))
ans1 <- my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 2L)

# You should get something like this
t(apply(ans1, 2, quantile, c(.025,.975)))
##                   2.5%      97.5%
## (Intercept) -0.1372435 0.05074397
## x            4.8680977 5.04539763
ans0
##                  2.5 %     97.5 %
## (Intercept) -0.1379033 0.04797344
## x            4.8650100 5.04883353


```
I did in fact get an answer very similar to the provided output, showing that the two are approximately the same. 

```{r}
system.time({my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 2L)})

system.time({my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 1L)})

```
We see that using 2 cores reduced the wait time by about 3 seconds during my first run, which for a large dataset would mean that code might only take days to run as opposed to weeks. 

## Problem 3: Short Answer

1. How you can decide if a problem is fit for  parallelization?

You can tell if a problem is fit for parallelization if it has many steps that can be completed without the output of another. For example, finding the max value of 1000 different dataframes are independent tasks that could be sped up through parallelization. Some task may require minor dependancies which can be addressed in the setup of the parallelization. These are still excellent candidates for parallelization. Tasks that are fully dependent on the task before cannot be parallelized. 

2. How you can think about the relationship between vectorization and parallelization? How do we know whether we should use one or the other or both?

Vectorization and parallelization are both methods used to speed up internal processes in computing. Vectorization is a way of running multiple processes at once, but only on one core. Instead of waiting on dependent processes (such as for loops), vectorization helps optimize the amount of wait time a computer needs to do. Parallelization is a method used to distribute multiple (hopefully unrelated) tasks onto multiple cores, which speeds up the process. All vectorization problems can be parallelized, but not all parallelization problems can be vectorized. To determine which to use (or both), one needs to look at the dependancy of the tasks, as well as the ability for the process to happen relatively simulataneously as opposed to linearly. 
